Parallel Adaptive Importance Sampling

### 2016-11-21 PAIS.tex
Added mixture model

\subsection{Sampling from Bimodal Distributions}

@@

\subsubsection{Target Distribution} \label{sec:tar}
We look at an `easy' problem, $B_1$, which has a KL divergence of
0.880, and a `harder' problem, $B_2$, which has a KL divergence of
3.647. Problem $B_1$ has two modes which are not too far apart. In
$B_2$ we increase the distance between the two modes which has the
effect of increasing the expected number of iterations that it takes
for a MCMC chain to jump between modes. These posteriors are shown in Figure~\ref{fig:problem 3 posteriors}.

@@

>>>>

\subsection{Sampling from Bimodal Distributions}\label{sec:bimodal}

@@

\subsubsection{Target Distribution} \label{sec:tar}

We look at an `easy' problem, $B_1$, which has a KL divergence of
0.880, and a `harder' problem, $B_2$, which has a KL divergence of
3.647. Problem $B_1$ has two modes which are not too far apart. In
$B_2$ we increase the distance between the two modes which has the
effect of increasing the expected number of iterations that it takes
for a MCMC chain to jump between modes. These posteriors are shown in Figure~\ref{fig:problem 3 posteriors}.

@@

\subsection{A Mixture Model}\label{sec:mixture}

The technique of mixture modelling employs well known parametric families to construct an approximating nonparametric distribution which may have a complex structure. Most commonly, Gaussian kernels are used since underlying properties in the data can often be assumed to follow a Gaussian distribution. An example would be if a practitioner were to measure the heights of one hundred adults, but failed to record their gender. The data could be considered as one population with two sub populations, male and female. The problem then might be to find the average height of adult males from the data. In this case, since height is often considered to follow a Gaussian distribution, it makes sense to model the population as a mixture of two univariate Gaussian distributions.

A well known problem in the Bayesian treatment of mixture modelling is that of identification, sometimes referred to as the label-switching phenomenon. The likelihood distribution for mixture models is invariant under permutations of the mixture labels. If a mixture has $n$ means and the point $(\mu_1, \dots, \mu_n)$ maximises the likelihood, then the likelihood will also be maximised by $(\mu_{\varphi(1)}, \dots, \mu_{\varphi(n)})$ for all permutations $\varphi(\cdot)$. This means that the number of modes in the posterior distribution is of order $\mathcal{O}(n!)$. As we have seen it can be hard for standard MCMC algorithms to obtain reliable inference for posterior distributions with a large number of modes, or even a small number of modes which are separated by a large distance.

\subsubsection{Target Distribution}\label{sec:mixture_target}

In particular we look at a Gaussian mixture where we assume that there are two subpopulations within the overall population. Since both subpopulations are assumed to follow a Gaussian distribution we have five parameters which we need to be estimated, two means $\mu_{1,2}$, two variances $\sigma^2_{1,2}$, and the probability, $p$, that an individual observation comes from one of the subpopulations. We have 100 data points, $D_i$, which we assume to be distributed according to
\[
	D_i \sim p\mathcal{N}(\mu_1, \sigma^2_1) + (1-p)\mathcal{N}(\mu_2, \sigma^2_2), \quad i = 1,\dots,100,
\]
where $p \in [0, 1]$, $\mu_{1,2} \in \mathbb{R}$ and $\sigma^2_{1,2} \in \mathbb{R}^+$. Due to the domains of these parameters, we assign the priors
\[
	p \sim \text{Beta}(1,1), \quad \mu_{1,2} \sim \mathcal{N}(0, 4) \quad \text{and} \quad \sigma^2_{1,2} \sim \text{Gamma}(\alpha=2, \beta=1).
\]
If we collect these parameters in the vector $\theta = (p, \mu_1, \sigma^2_1, \mu_2, \sigma^2_2)^\top$, the resulting posterior distribution is
\begin{equation}\label{eq:mixture_posterior}
	\pi(\theta|D) \propto \prod\limits_{i=1}^{100} (p\mathcal{N}(D_i;\mu_1, \sigma^2_1) + (1-p)\mathcal{N}(D_i;\mu_2, \sigma^2_2))\prod\limits_{i=1}^5 \pi_0^i(\theta_i),
\end{equation}
where $\pi_0^i(\cdot)$ is the prior density function corresponding to $\theta_i$.

\begin{figure}[htb]
\centering
\includegraphics[width=\textwidth]{"figures/PAIS_best_posterior"}
\caption{The posterior distribution of $\theta$ as given in Equation~\eqref{eq:mixture_posterior}. The main diagonal contains the marginal distributions of each $\theta_i$, and the lower triangular contours represents the correlation between pairs of parameters.}
\label{fig:mixture_posterior}
\end{figure}

Figure \ref{fig:mixture_posterior} presents a visualisation of the posterior distribution for this problem, created from a single PAIS simulation.

\subsubsection{Implementation}\label{sec:mixture_implementation}

We have no analytic form for the posterior distribution for this model, and MH cannot reliably sample from this type of bimodal distribution. We do however know that the probability density should be evenly divided between the two modes. This is due to the symmetric prior for $p$, and and the same priors being assigned to $\mu_1$ and $\mu_2$, and also to $\sigma^2_1$ and $\sigma^2_2$. To decide which mode a sample belongs to we define a plane which bisects the posterior so that each point on this plane lies exactly halfway between the two true solutions to the inverse problem i.e. the value of $\theta$ used to generate the data, and also the $\theta$ obtained by a relabelling of the parameters. Now that we can assign a sample to a particular mode, we can calculate the density in each mode by summing the weights associated to all samples in that mode,
\[
	\bar{w}_k = \sum\limits_{i=1}^N w_i\text{I}(X_i \in \text{Mode $k$}), \quad k = 1, 2,
\]
and the relative error in the amount of density in each mode is then
\[
	w_\text{error} = 2\left|\frac{\bar{w}_1}{\bar{w}_1+\bar{w}_2} - \frac{1}{2}\right|.
\]

Since the probability $p$ is constrained to lie in the interval $[0,1]$, and the variances must be positive, it can be wasteful to use Gaussian proposal distributions. For this example we make proposals by matching the proposal distribution to the prior, scaling the variance and centring the proposal at the previous value. So
\[
	q_p \sim \text{Beta}(\delta^{-2}p, \delta^{-2}(1-p)), \quad q_{\mu_{1,2}} \sim \mathcal{N}(\mu_{1,2}, 4\delta^2) \quad \text{and} \quad q_{\sigma^2_{1,2}} \sim \text{Gamma}(\alpha^*, \beta^*),
\]
where $\alpha^* = \sigma^2_{1,2}\beta^*$, $\beta^* = \sigma^2_{1,2}/2\delta^2$ and $\delta$ is a scaling parameter to be tuned. This means that our proposal distributions will not be a mixture of multivariate Gaussians, but independent mixtures of univariate Beta, Gamma and Gaussian distributions.

In the numerics which follow we have increased the ensemble size from $M=50$ to $M=500$ to compensate for the increase in dimension. We also use the AMR algorithm for the resampling step because of the reduced computational cost. The method otherwise remains the same as in previous examples. We perform test runs to find the optimal scaling parameters considering convergence to modes with equal density. We then calculate the convergence rates of the algorithms by producing 50 million samples from the posterior with each algorithm, and repeat the simulation 32 times.

% numerics.
\subsubsection{Convergence of MH vs PAIS}\label{sec:mixture_conv}

\begin{table}[!htb]
      \centering
        \begin{tabular}{|l|r|r|}
	\hline
	Algorithm	& MH & PAIS \\ \hline
	$\delta^*$	& 1.3e-1     & 2.3e-1 \\
	\hline
	\end{tabular}
	\vspace{1mm}
	\caption{Optimal values of the scaling parameter. The MH algorithm is optimised using the acceptance rate, and the PAIS algorithm is optimised using the effective sample size.}
	\label{table:mixture_opt_beta}
\end{table}

The optimal scaling parameters for the MH and PAIS algorithms with the proposal distributions described in Section~\ref{sec:mixture_implementation} are given in Table~\ref{table:mixture_opt_beta}.

\begin{figure}[htb]
\centering
\includegraphics[width=0.8\textwidth]{"figures/Mode_proportions"}
\caption{Convergence of the PAIS algorithm for the mixture model described in Section~\ref{sec:mixture}. Implementation described in Sections~\ref{sec:mixture_implementation} and \ref{sec:mixture_conv}. Resampling is performed using the AMR scheme.}
\label{fig:mixture_modes}
\end{figure}

Convergence of the relative error for the two algorithms is displayed in Figure~\ref{fig:mixture_modes}. PAIS converges at the expected $\mathcal{O}(1/\sqrt{N})$ rate, whereas the MH algorithm converges to locally smooth histograms but with the wrong proportion of samples in each mode. The relatively low value of the error for the MH example is due to the priors covering the sample space evenly, however since transitions are near impossible with a small value of the scaling parameter, this error will take a very long time to reduce. This problem was discussed in Section~\ref{sec:bimodal}.
